---
title: "Working with matrices in R"
author: "Adapted by Milan Malfait"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
    html_document:
      code_download: true
      theme: cosmo
      toc: true
      toc_float: true
      highlight: tango
      number_sections: false
---

```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

***

## Creating and manipulating matrices

We will create a $4 \times 2$ matrix, starting from a vector of values.

```{r create_matrix}
## Define the matrix X, filling matrix "byrow": 4 x 2
X <- matrix(c(3, 6, 1, 2, 0.23, 0.46, -2, -4), nrow = 4, byrow = TRUE)
X

## Alternative way to define the matrix, `byrow = FALSE` is the default
X <- matrix(c(3, 1, 0.23, -2, 6, 2, 0.46, -4), ncol = 2, byrow = FALSE)
X

# Creating a second matrix: 2 x 4, using the square of `X`
Y <- matrix(X^2, nrow = 2, ncol = 4)
Y

## Check that we really have a matrix
class(X)
```
Generally you can check the arguments of a function, e.g. `matrix`, in R with `?matrix`. 

The `dim` function tells you the __dimensions__ of the matrix, `nrow` and `ncol` give you the number of rows and columns, respectively.

```{r}
dim(X)  # nr. of rows, nr. of columns
nrow(X)
ncol(X)

dim(Y)
```


### Subsetting

Note that R sees a matrix as `X[row, column]`, so you can subset `X` with

```{r subset_matrix}
X[1, ] # first row of X
X[, 2] # second column of X
X[1:2, ] # first 2 rows of X
```

Note that, by default, subsetting a single row or column from a matrix __returns a vector__, not a column- or row-matrix as might be expected.
We can see this from the fact that the subset does not have dimensions anymore (an attribute only defined for matrices in R):

```{r}
dim(X[, 1]) # NULL = non-existent
class(X[, 1]) # not matrix but numeric vector
```

This can lead to unexpected behavior when doing __matrix multiplication__ (see further below).
To avoid this, we can set `drop = FALSE` in the subsetting brackets.

```{r}
## This creates a "column-vector"
X[, 1, drop = FALSE]
dim(X[, 1, drop = FALSE])
```


### Row- and column-binding

You can also create matrices by 'binding' together vectors (of the same length) column-wise or row-wise with `cbind()` and `rbind()`, respectively.

```{r}
# Create vectors from the rows of X
(x1 <- X[1, ])
(x2 <- X[2, ])
(x3 <- X[3, ])
(x4 <- X[4, ])

# Bind them back together row-wise: re-creates X
rbind(x1, x2, x3, x4)

# Bind them back together column-wise: (essentially the transpose of X)
cbind(x1, x2, x3, x4)
```

Binding together 2 matrices also works (but be mindful of their dimensions!).

```{r, error=TRUE}
# Row-binding two matrices
# X and Y don't have the same dimensions (4x2 vs. 2x4),
# so binding them directly doesn't work
rbind(X, Y)

# For rbind we need the same number of columns
rbind(X, Y[, 1:2]) # works

# For cbind, same number of rows
cbind(X[1:2, ], Y)
```

Note that R automatically creates row names or column names from the variable names when using `rbind()` and `cbind()`.

To add row and columns names to an existing matrix, use `rownames()`, `colnames()` or the more general `dimnames()`.

```{r}
rownames(X) <- c("row_1", "row_2", "row_3", "row_4")
colnames(X) <- c("col_1", "col_2")
X

## Using `NULL` removes the dimnames again
dimnames(X) <- NULL
X
```


### Diagonal elements and diagonal matrices: `diag()`

```{r}
## Get diagonal elements from X, returns vector
diag(X)

## Create diagonal matrix, vector as input, returns matrix
diag(c(1, 2, 3 ,4))
```

### Identity matrices

The `diag` function can also be used to create identity matrices with specific dimensions

```{r}
diag(1, nrow = 4, ncol = 4)  # 4 x 4
diag(1, nrow = 2, ncol = 4)  # 2 x 4
diag(1, nrow = 4, ncol = 2)  # 4 x 2
```



## Math operations

### Element-wise operations

#### Scalars and vectors

Scalar operations occur element-wise for matrices.

```{r}
X # for reference
X + 1
X - 100
2 * X
X / 3
X^2
```

Note that `^2` also occurs element-wise, *not* using matrix multiplication (see further below)!

You can also do operations with vectors, though you should be careful with the dimensions.
`R` will add the vector to each __column__ of the matrix.
When using a vector with a different length than the number of rows (elements in a column), the vector elements will be *recycled*.
When the longer object is not a multiple of the shorter, the operation still occurs but a *warning* is produced (technically a `message`).

```{r}
X
v4 <- c(1, 2, 3, 4)  # vector of length 4
X + v4

v2 <- c(1, 2)  # vector of length 2, will be recycled
X + v2

v3 <- c(1, 2, 3) # vector or length 3, also recycled with warning
X + v3

v5 <- c(1, 2, 3, 4, 5) # vector or length 5, also recycled with warning
X + v5
```

#### Two matrices

Define a second matrix `Y` with same dimensions as `X`.

```{r}
# We will generate a matrix Y with 2s and 4s at random places (using `sample`)
# Setting a seed allows reproducibility
set.seed(1)
Y <- matrix(sample(c(2, 4), size = nrow(X) * ncol(X), replace = TRUE), 
            nrow = nrow(X), ncol = ncol(X))
Y
```

Using the standard math operators occurs element-wise.

```{r}
X + Y
X - Y
X * Y
X / Y
X ^ Y
```


### Matrix algebra

<https://www.statmethods.net/advstats/matrix.html>

#### Transpose: `t()`

$\mathbf{X}^T$

```{r}
t(X)
```

#### Matrix multiplication: `%*%`

Note that `X` and `Y` have the same dimensions, so multiplying them straight away doesn't work (incompatible dimensions)

```{r, error=TRUE}
## This produces an error because the dimensions are not compatible
X %*% Y
```

Instead, we can transpose `X` and compute $\mathbf{X}^T \cdot \mathbf{Y}$:

```{r}
t(X) %*% Y

## `crossprod()` is a shortcut for this operation, and is slightly faster
crossprod(X, Y)

## `tcrossprod()` calculates X.Y^T
X %*% t(Y)
tcrossprod(X, Y)
```

The `crossprod()` and `tcrossprod()` functions also work with a single matrix as input, in which case they calculate $\mathbf{X}^T\mathbf{X}$ and $\mathbf{X}\mathbf{X}^T$, respectively:

```{r}
## X^TX
crossprod(X)
## XX^T
tcrossprod(X)
```

As already mentioned before, subsetting a matrix with `[` returns a vector by default.
Doing matrix multiplication with vectors in R can be ambiguous as R will "guess" itself whether the vector should be treated as a column- or row-vector.
See `` ?`%*%` `` for details.

The example below illustrates some potentially unexpected results

```{r, error=TRUE}
## Selecting the first column of X, we would expect this to be treated as a column
## vector with dimensions 4 x 1...
X[, 1]
dim(X[, 1])
## ...but in R vectors don't have dimensions

## Mathematically, multiplicating the first column of X with X is not possible
## (multiplying a 4x1 with a 4x2 matrix doesn't work)
## So we would expect the following code to throw an error...
X[, 1] %*% X
## ...but it doesn't and instead returns the result as if X[, 1] is a row-vector
```

To avoid confusion and when you intend to use single rows or vectors from a matrix as they were single-row or single-column matrices, it's better to use `drop = FALSE` when subsetting.

```{r, error=TRUE}
X1 <- X[, 1, drop = FALSE]
## Now X1 does have dimensions
dim(X1)

## And the following code throws an error
X1 %*% X
```

You can also do matrix multiplication with 2 vectors, in which case again R will guess if they should be treated as column- or row-vectors.

```{r}
(a <- seq_len(3))
(b <- a + 1)
a %*% b
```


#### Inverse: `solve()`

__Inverse__: $\mathbf{X}^{-1}$ (only for square matrices)

```{r}
## Create new square matrix Z
(Z <- matrix(1:4, nrow = 2, ncol = 2))

## Compute inverse
(Z_inv <- solve(Z))

## Check that it's indeed the inverse
Z %*% Z_inv  # the 2x2 identity matrix
```

__Warning:__ DON'T USE `X^-1` to compute a matrix inverse!

The notation might be confusing because of the way we mathematically describe the inverse of a matrix ($\mathbf{X}^{-1}$).
In R, however, `X^-1` computes __element-wise__ inverses of each value.

```{r}
Z^-1
```

If you're wondering why the name "`solve`" is used for the function to compute a matrix inverse, it's because `solve` can do more than just that.
See `?solve` for details.


## Rank of a matrix

Calculating the rank of the matrix $\mathbf X$.
There are several ways to do this in R.
We'll use the `qr()` function (see `?qr`).
An alternative would be to use `rankMatrix` from the [*Matrix*](https://cran.r-project.org/package=Matrix) package: `Matrix::rankMatrix()`.

```{r rank}
## Run `qr` and access the `rank` from the results with `$`
rank <- qr(X)$rank
rank
```

We see that the rank of `X` is lower than its number of columns and rows, meaning the matrix is [*not* of full rank](https://en.wikipedia.org/wiki/Rank_(linear_algebra)#Main_definitions).

Considering the contents of `X`, can you see why this is?

__Hint__: 

```{r}
all(X[, 2] == 2 * X[, 1])
```



## Session info {-}

<details><summary>Session info</summary>

```{r session_info, echo=FALSE, cache=FALSE}
Sys.time()
sessioninfo::session_info()
```

</details>

## [Home](https://statomics.github.io/HDA2020/) {-}
