---
title: "Analysis of High Dimensional Data - Lab 3"
subtitle: "Penalized regression techniques for high-dimensional data"
author: "Adapted by Milan Malfait"
date: "05 Nov 2020"
---

```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  out.width = "100%"
)
options(
  warnPartialMatchDollar = FALSE,
  warnPartialMatchAttr = FALSE,
  warnPartialMatchArgs = FALSE
)
```

***

```{r libraries, warning=FALSE, message=FALSE}
## install packages with:
## install.packages(c("glmnet", "pls", "NormalBetaPrime", "pROC", "boot"))
library(NormalBetaPrime)
library(glmnet)
library(pls)
library(pROC)
library(boot)
```


# Introduction

**In this lab session we will look at the following topics**

  - Demonstrate why low dimensional prediction modeling fails in high dimension.
  - Carry out Principal Component Regression (PCR)
  - Use `glmnet()` to carry out ridge regression, lasso and elastic net
  - Evaluation of these prediction models
  

## The dataset

In this practical, we will use the dataset `eyedata` provided by
the [__NormalBetaPrime__ package](https://cran.r-project.org/web/packages/NormalBetaPrime/index.html).
This dataset contains gene expression data of 200
genes for 120 samples. The data originates from microarray experiments
of mammalian eye tissue samples.

The dataset consists of two objects:

  - `genes`: a $120 \times 200$ matrix with the expression levels of 200 genes
  (columns) for 120 samples (rows)
  - `trim32`: a vector with 120 expression levels of the TRIM32 gene.


```{r load-data}
data(eyedata)
ls() # List objects in current R session
```

The goal of this exercise is to predict the expression levels of
TRIM32 from the expression levels of the 200 genes measured in the
microarray experiment. For this, it makes sense to start by constructing
centered (and possibly scaled) data. We store this in two matrices
`X` and `Y`:

```{r prepare-data}
X <- scale(genes, center = TRUE, scale = TRUE) 
Y <- scale(trim32, center = TRUE)
```

Remember that scaling avoids that differences in levels of magnitude
will give one variable (gene) more influence in the result. This has
been illustrated in the [second practical session](./Lab2-PCA.html) as well.
For the `Y` vector, this is less of an issue as we're talking about a single variable.
Not scaling will make the predictions interpretable as "deviations from the
mean".

## The curse of singularity

We begin by assuming that the predictors and the outcome have been
centered so that the intercept is 0.
We are presented with the usual regression model:

$$
Y_i=\beta_i X_{i1}+\dots+\beta_pX_{ip}+\epsilon_i \\ 
\text{ Or } \mathbf{Y}={\mathbf{X}}{\boldsymbol{\beta}} +{\boldsymbol{\epsilon}}
$$

Our goal is to get the least squares estimator of
${\boldsymbol{\beta}}$, given by

$$
\hat{{\boldsymbol{\beta}}}= (\mathbf{X}^T{\mathbf{X}})^{-1}{\mathbf{X}}^T{\mathbf{Y}}
$$

in which the $p \times p$ matrix
$({\mathbf{X}}^T{\mathbf{X}})^{-1}$ is crucial!
To be able to calculate the inverse of ${\mathbf{X}}^T \mathbf{X}$,
it has to be of full rank $p$, which would be 200 in this case.
Let's check this:

```{r singularity-problem, error=TRUE}
dim(X) # 120 x 200, so p > n!
qr(X)$rank

XtX <- crossprod(X) # calculates t(X) %*% X more efficiently
qr(XtX)$rank

# Try to invert using solve: 
solve(XtX)
```

We realize we cannot compute
$({\mathbf{X}}^T{\mathbf{X}})^{-1}$ because the rank of
$({\mathbf{X}}^T{\mathbf{X}})$ is less than $p$ hence we can’t
get $\hat{{\boldsymbol{\beta}}}$ by means of least squares! 
This is generally referred to as the __[singularity](https://www.statistics.com/glossary/singularity/) problem__.


# Principal component regression

A first way to deal with this singularity, is to bypass it using principal components.
Since $\min(n,p) = n = 120$, 
PCA will give `r min(dim(X))` components, each being a linear combination of the 
$p$ = `r ncol(X)` variables.
These `r min(dim(X))` PCs contain all information present in the original data.
We could as well use an approximation of ${\mathbf{X}}$, i.e using just a few ($k<120$) PCs.
So we use PCA as a method for reducing the dimensions while retaining
as much variation between the observations as possible.
Once we have these PCs, we can use them as variables in a linear regression model.

## Classic linear regression on PCs

We first compute the PCA on the data with `prcomp`.
We will use an arbitrary cutoff of $k = 4$ PCs to illustrate the process of performing regression on the PCs.

```{r PC-regression}
k <- 4 # Arbitrarily chosen k=4
pca <- prcomp(X)
Vk <- pca$rotation[, 1:k] # the loadings matrix
Zk <- pca$x[, 1:k] # the scores matrix

# Use the scores in classic linear regression
pcr_model1 <- lm(Y ~ Zk)
summary(pcr_model1)
```

As $\mathbf{X}$ and $\mathbf{Y}$ are centered, the intercept is 
approximately 0.

The output shows that PC1 and PC4 have a $\beta$ estimate that 
differs significantly from 0 (at $p < 0.05$), but the results can't be readily 
interpreted, since we have no immediate interpretation of the PCs.


## Using a binary response/outcome variable

Instead of predicting the actual gene expression levels, researchers are
often more interested in whether or not an expression level crosses a
certain threshold, indicating that the gene has been turned on.
We can easily construct such a dataset for the TRIM32 outcome.
We define "success" (1) as having a value above the median, and "failure" (0) as a value below the median.
Discretizing a continuous variable like this is also sometimes called __dichotomization__.

```{r binarize-Y}
Ybin <- ifelse( Y > median(Y), 1, 0)
table(Ybin)
```

The `ifelse()` function is a vectorized function that translates a 
logical vector into another one based on its values. See also
`?ifelse` for more information.

Note that now we cannot use an ordinary linear regression model anymore, because our response variable is no longer (approximately) normal (it can only take on values 0 and 1!).
Instead, we will use __logistic regression__, which models the response variable $Y$ with a *binomial* distribution and uses a *logit* link-function between the response and the linear predictor.
Logistic regression belongs to the class of __Generalized Linear Models (GLM)__ (in fact, standard linear regression is also a special case of GLM) and is implemented in R in the `glm` function.
The type of GLM fitted by `glm` is controlled with the `family` argument.
For logistic regression, we specify `family = "binomial"` (by default, this argument is set to `"gaussian"`, which is equivalent to `lm`, see `?glm` for details).

```{r pc-logistic-regression}
pcr_logistic_model <- glm(Ybin ~ Zk, family = "binomial")
summary(pcr_logistic_model)
```


## Using the package `pls`

PCR can also be performed using the `pcr()` function from the
package *[pls](https://CRAN.R-project.org/package=pls)*
__directly on the data__ (so without having to first perform the PCA manually).
When using this function, you have to keep a few things in mind:

  1. the number of components (PCs) to use is passed with the argument `ncomp`
  2. the function allows you to scale (set `scale = TRUE`) and
  center (set `center = TRUE`) the predictors first (in the example here, $\mathbf{X}$ has already been centered and scaled).
  
You can use the function `pcr()` in much the same way as you would
use `lm()`. The resulting fit can easily be examined using the 
function `summary()`, but the output looks quite different from
what you would get from `lm`.

```{r PC-regression-pls-package}
# X is already scaled and centered, so that's not needed.
pcr_model2 <- pcr(Y ~ X, ncomp = 4)
summary(pcr_model2)
```

First of all the output shows you the data dimensions and the fitting
method used. In this case, that is PC calculation based on SVD. The
`summary()` function also provides the percentage of variance
explained in the predictors and in the response using different numbers
of components. For example, the first PC only captures 61.22% of all
the variance, or information in the predictors and it explains 62.9%
of the variance in the outcome. Note that for both methods the choice of
the number of principal components was arbitrary chosen to be 4.

At a later stage, we will look at how to choose the number of components
that has the __smallest prediction error__.


# Ridges, Lassos and Elastic Nets {#elnet-theory}

Ridge regression, lasso regression and elastic nets are all closely
related techniques, based on the same idea: add a penalty term to 
the estimating function so $({\mathbf{X}}^T{\mathbf{X}})$
becomes full rank again and is invertible. Two different penalty 
terms or regularization methods can be used:

1. L1 regularization: this regularization adds a term ${\gamma_1\|\boldsymbol{\beta}\|_{1}}$ to the estimating equation.
The term will add a penalty based on the *absolute value* of the
magnitude of the coefficients. This is used by the __lasso regression__
 
$$
 \hat{\boldsymbol{\beta}}^{\text{lasso}} = \text{argmin}_{\boldsymbol{\beta}}\displaystyle({(\mathbf{Y}-\mathbf{X}\boldsymbol{\beta})^T(\mathbf{Y}-\mathbf{X}\boldsymbol{\beta})+{\gamma_1\|\boldsymbol{\beta}\|_{1}}}\displaystyle)
$$

2. L2 regularization: this regularization adds a term ${\gamma_2\|\boldsymbol{\beta}\|_{2}^{2}}$ to the estimating equation.
The penalty term is based on the square of the magnitude of the 
coefficients. This is used by __ridge regression__.

$$
 \hat{\boldsymbol{\beta}}^{\text{ridge}} = \text{argmin}_{\boldsymbol{\beta}}\displaystyle({(\mathbf{Y}-\mathbf{X}\boldsymbol{\beta})^T(\mathbf{Y}-\mathbf{X}\boldsymbol{\beta})+{\gamma_2\|\boldsymbol{\beta}\|_{2}^{2}}}\displaystyle)
$$

Elastic nets combine both types of regularizations. It does so by 
introducing a $\alpha$ mixing parameter that essentially combines
the L1 and L2 norms in a weighted average.

$$
 \hat{\boldsymbol{\beta}}^{\text{el.net}} = \text{argmin}_{\boldsymbol{\beta}}\displaystyle({(\mathbf{Y}-\mathbf{X}\boldsymbol{\beta})^{T}(\mathbf{Y}-\mathbf{X}\boldsymbol{\beta})+{\alpha \gamma_1\|\boldsymbol{\beta}\|_{1}}+ {(1 - \alpha)\gamma_2\|\boldsymbol{\beta}\|_{2}^{2}}}\displaystyle)
$$



# Exercise: Verification of ridge regression

In least square regression the minimization of the estimation function
$|{\mathbf{Y} - \mathbf{X} \boldsymbol{\beta}}\|^{2}_{2}$ leads to the solution ${\boldsymbol{\hat{\beta}}=(\mathbf{X^TX})^{-1}\mathbf{X^TY}}$. 

For the penalized least squares criterion used by ridge regression, you minimize 
$\|{\mathbf{Y}-\mathbf{X}\boldsymbol{\beta}\|^{2}_{2}}+\gamma{\boldsymbol{\|\beta\|^{2}_{2}}}$
which leads to following solution:

$$
{\boldsymbol{\hat{\beta}}=(\mathbf{X^TX}}+\gamma{\mathbf{I}})^{-1}{\mathbf{X^TY}}
$$

where $\mathbf{I}$ is the $p \times p$ identity matrix.

The ridge parameter $\gamma$ *shrinks* the coefficients towards 0, with $\gamma = 0$ being equivalent to OLS (no shrinkage) and $\gamma = +\infty$ being equivalent to setting all $\hat{\beta}$'s to 0.
The optimal parameter lies somewhere in between and needs to be tuned by the user.


## Tasks {-}

Solve the following exercises using R.

#### 1. Verify that ${\mathbf{(X^TX}}+\gamma{\mathbf{I}})$ has rank $200$, for any $\gamma>0$ of your choice. {-}

<details><summary>Solution</summary>
```{r}
XtX <- crossprod(X)
p <- ncol(X)
gamma <- 2 # My choice

# Compute penalized matrix
XtX_gammaI <- XtX + (gamma * diag(p))
dim(XtX_gammaI)
qr(XtX_gammaI)$rank == 200 # indeed
```
</details>


#### 2. Check that the inverse of ${\mathbf{(X^TX}}+\gamma{\mathbf{I}})$ can be computed. {-}

<details><summary>Solution</summary>
```{r}
# Yes, it can be computed (no error)
XtX_gammaI_inv <- solve(XtX_gammaI)
str(XtX_gammaI_inv)
```
</details>


#### 3. Finally, compute ${\boldsymbol{\hat{\beta}}=(\mathbf{X^TX}}+\gamma{\mathbf{I}})^{-1}{\mathbf{X^TY}}$. {-}

<details><summary>Solution</summary>
```{r ridge-beta-estimates}
## Calculate ridge beta estimates
## Use `drop` to drop dimensions and create vector
ridge_betas <- drop(XtX_gammaI_inv %*% t(X) %*% Y)
length(ridge_betas) # one for every gene
summary(ridge_betas)
```

We have now manually calculated the ridge regression estimates.

</details>


<!-- If we perform SVD on ${\mathbf{X}}$ we get -->
<!-- ${\mathbf{X}}={\mathbf{UDV^T}}={\mathbf{ZV^T}}$. -->
<!-- Replacing ${\mathbf{X}}$ with ${\mathbf{ZV^T}}$ gives -->
<!-- ${\boldsymbol{\hat{\beta}}=\mathbf{V(Z^TZ}}+\gamma{\mathbf{I}})^{-1}{\mathbf{Z^TY}}$ -->

<!-- #### 4. Check to see that ${\mathbf{(X^TX}}+\gamma{\mathbf{I}})$ has a higher dimension than ${\mathbf{(Z^TZ}}+\gamma{\mathbf{I}})$. {-} -->

<!-- And therefore inverting ${\mathbf{(Z^TZ}}+\gamma{\mathbf{I}})$ is much faster than inverting -->
<!-- ${\mathbf{(X^TX}}+\gamma{\mathbf{I}})$ (You don’t have to check this). -->

<!-- <details><summary>Solution</summary> -->
<!-- ```{r} -->
<!-- X.svd<-  svd(X) -->
<!-- Uk <- X.svd$u -->
<!-- Dk <- diag(X.svd$d) -->
<!-- Zk <- Uk%*%Dk -->
<!-- Vk <- X.svd$v -->
<!-- p<-dim(Zk)[1] -->
<!-- tZZ_gammaI<-(t(Zk)%*%Zk)+gamma*diag(p) -->
<!-- dim(tZZ_gammaI) -->

<!-- # Calculate the beta -->
<!-- hatbeta_svd <- Vk %*% solve(tZZ_gammaI) %*% t(Zk) %*% Y -->

<!-- # Check whether the same -->
<!-- comparison <- cbind(hatbeta, hatbeta_svd) -->
<!-- head(comparison) # to not show everything  -->
<!-- ``` -->
<!-- </details> -->

<!-- We therefore conclude that -->
<!-- ${\boldsymbol{\hat{\beta}}=\mathbf{V(Z^TZ}}+\gamma{\mathbf{I}})^{-1}{\mathbf{Z^TY}}$ -->
<!-- is a faster way for computing ${\boldsymbol{\hat{\beta}}}$. -->



# Performing ridge and lasso regression with `glmnet`

The package *[glmnet](https://CRAN.R-project.org/package=glmnet)* provides a
function `glmnet()` that allows you to fit all three types of regressions. Which
type is used, can be determined by specifying the `alpha` argument. For a
__ridge regression__, you set `alpha` to 0, and for a __lasso regression__ you
set `alpha` to 1. Other `alpha` values between 0 and 1 will fit a form of
elastic net. This function has slightly different syntax from the other
model-fitting functions. To be able to use it, you have to pass a `x` matrix as
well as a `y` vector, and you don't use the formula syntax.

The gamma value, which controls the "strength" of the penalty, can be passed by
the argument `lambda` (notation isn't always consistent between text books and
software...). The function `glmnet()` can also carry out a search for finding
the best gamma value for a fit. This can be done by passing multiple values to
the argument `lambda`. If not supplied, `glmnet` will generate a range of values
itself, based on the data whereby the number of values can be controlled with
the `nlambda` argument. This is generally the recommended way to use `glmnet`,
see `?glmnet` for details.

For a thorough introduction to the __glmnet__ package and elastic net models in
general, see the
[glmnet introduction vignette](https://cran.r-project.org/web/packages/glmnet/vignettes/glmnet.pdf)


## Ridge regression

Let's perform a ridge regression in order to predict expression levels
of the TRIM32 gene using the 200 gene probes data. We can start by
using a $\gamma$ value of 2.

```{r glmnet-ridge-regression}
gamma <- 2
ridge_model <- glmnet(X, Y, alpha = 0, lambda = gamma)

# have a look at the first 10 coefficients
coef(ridge_model)[1:10]
```

The first coefficient is the intercept, and is again essentially 0. But
a value of 2 for $\gamma$ might not be the best choice, so let's see how
the coefficients change with different values for $\gamma$.

We will create a *grid* of $\gamma$ values, i.e. a range of values that will be
used as input for the `glmnet` function. Note that this function can take a
vector of values as input for the `lambda` argument, allowing to fit multiple
models with the same input data but different hyperparameters.

```{r ridge-regression-grid-search}
grid <- seq(1, 10000, length = 100)
ridge_mod_grid <- glmnet(X, Y, alpha = 0, lambda = grid)

# Plot the coefficients against the (natural) LOG lambda sequence!
# see ?plot.glmnet
plot(ridge_mod_grid, xvar = "lambda", xlab = "log(gamma)")
# add a vertical line at gamma = 2
text(log(gamma), -0.05, labels = expression(gamma == 2), 
     adj = -0.5, col = "firebrick")
abline(v = log(gamma), col = "firebrick", lwd = 2)
```

This plot is known as a __coefficient profile plot__, each colored line
represents a coefficient $\hat{\beta}$ from the regression model and shows how
they change with increased values of $\gamma$ (on the log-scale)
^[Note: `log()` in R is the __natural logarithm__ by default (base $e$) and we
will also use this notation in the text (like the x-axis title on the plot above).
This might be different from the notation that you're used to ($\ln()$).
To take logarithms with a different base in R you can specify the `base = `
argument of `log` or use the shorthand functions `log10(x)` and `log2(x)` for
base 10 and 2, respectively].

Note that for higher values $\gamma$, the coefficient estimates become closer to 0,
showing the *shrinkage* effect of the ridge penalty.

Similar to the PC regression example, we chose $\gamma=2$ and the grid rather
arbitrarily. We will see subsequently, how to choose $\gamma$ that minimizes the
prediction error.


## Lasso regression

Lasso regression is also a form of penalized regression, but we do not have any
analytic solution of $\hat{{\boldsymbol{\beta}}}$ as in least squares
and ridge regression. In order to fit a lasso model, we once again use
the `glmnet()` function. However, this time we use the argument
`alpha = 1`

> Verify that this indeed corresponds to lasso regression using the equations
from [Section 3](#elnet-theory).

Other than this, we proceed just as we did when fitting a ridge regression model.
However, we will not provide a custom sequence of $\gamma$ (`lambda`) values here
but instead rely on `glmnet`'s 

```{r glmnet-lasso-regression}
# Note that the glmnet() function can supply gamma automatically
lasso_model <- glmnet(X, Y, alpha = 1)
plot(lasso_model, xvar = "lambda", xlab = "log(gamma)")
```


# Exercise: performing penalized regressions with `glmnet`

## Tasks {-}

#### 1. Use the grid of $\gamma$s and fit a ridge regression model with the binary outcome variable `Ybin` and predictors `X`. Hint: in the `glmnet()` function use `family="binomial"`. {-}
    
```{r}
#Binary version
library(glmnet)
grid <-seq(1,1000, length=100)# arbitrary grid

ridge.mod <- glmnet(X,Ybin,alpha=0,lambda=grid,family = "binomial")
```

#### 2. Fit a Lasso model with the binary outcome variable `Ybin` and predictors `X`. Again you use the binomial family {-}
    
```{r}
lasso.mod <- glmnet(X,Ybin, alpha = 1,
                    family = "binomial")
```

#### 3. Plot your fitted models. {-}

```{r, fig.show='hold', out.width='49%'}
plot(ridge.mod, xvar ="lambda", xlab ="gamma")

plot(lasso.mod,xvar ="lambda", xlab ="gamma")
```

Again we ask the question which $\gamma$ is the right one? Generally the
number of PCs in PC regression and $\gamma$ in the ridge and Lasso
regression are called tuning parameters. The different number of
PCs/$\gamma$, corresponds to different models which can be represented
as a class of models $\mathcal{M}$. We now look at how to chose turning
parameters and hence optimal prediction models from $\mathcal{M}$.


# Evaluation of prediction models

Before we begin, we use the `set.seed()` function in order to set a seed
for R’s random number generator, so that we will all obtain precisely
the same results as those shown below. It is generally good practice to
set a random seed when performing an analysis such as cross-validation
that contains an element of randomness, so that the results obtained can
be reproduced precisely at a later time.

We begin by using the `sample()` function to split the set of samples
into two subsets, by selecting a random subset of 80 observations out of
the original 120 observations. We refer to these observations as the
**training** set.

```{r create training set}
set.seed(1)
# Select 80 out of 120 (row)numbers
trainID <- sample(120 ,80)

#Training data
trainX <- X[trainID,] 
trainY <- Y[trainID]

#Test data
testX <- X[-trainID,] 
testY <- Y[-trainID]
```

Here we use a shortcut in the sample command; see `?sample` for details.

We are interested in the **extra-sample** error of our models. The
extra-sample error will be estimated using cross-validation (CV). We
will see both LOOCV and CV$_k$ ( k-folds cross validation).

-   For the continuous outcome we will use the mean squared error (MSE).

-   For the binary outcome, we will use the area under the receiver
    operating characteristic curve (ROC), the AUC and the
    misclassification error.

-   We use the test/validation dataset for final model evaluation. No
    estimation of model parameters whatsoever is done on the test
    dataset.

The LOOCV and CV$_k$ estimates can be automatically computed for any
generalized linear model using the `glm()` and `cv.glm()` functions. The
`cv.glm()` function is part of the `boot` package.  

Let’s begins by creating our cost/evaluation measure functions. 
For the AUC function, we simply use the function `auc` of the `pROC` package.



```{r functions to calculate MSE, AUC and misclassification}
suppressMessages(library(pROC))

MSE <- function(obs, pred){
  mse <- mean((obs - pred)^2)
  return(mse)
}

MISERR <- function(obs, pred, cutoff = 0.5){
  ypred <- as.numeric(pred > cutoff)  # translates TRUE/FALSE to 1/0
  tab <- table(obs, ypred)
  miserr <- 1 - sum(diag(tab))/sum(tab)
  return(miserr)
}
```

## PC regression

We start with the PC regression and choose the number of PCs that minimizes
a chosen extra-sample error. We use the function `predict()` to calculate
the prediction errors. As our training dataset contains now only
80 rows, we can test for a maximum of 79 PCs.

```{r PC regression crossvalidation}
suppressMessages(library(boot)) # for the cv.glm function

max.n.comps <- 79
cv.pcr.errors <- rep(NA, max.n.comps)

for(i in 1:max.n.comps){
  # Train the model
  pcr.train <- pcr(trainY ~ trainX, ncomp = i)
  # Calculate the predictions
  ypred <- predict(pcr.train, newdata = testX)
  # Calculate the extra-sample error
  cv.pcr.errors[i] <- MSE(testY,ypred)
}

```

Now we have a vector `cv.pcr.errors` that contains the MSE for every
number of components. We can use this information to create a plot,
and to find the number of components with the lowest MSE value:

```{r find optimal number of components}
nPC.at.min <- which.min(cv.pcr.errors)
plot(1:max.n.comps, cv.pcr.errors, type = "l",
     ylab = "MSE", xlab = "number PCs")
abline(v = nPC.at.min, col = "red")
```

This is a manual procedure, but we can also use a K-fold crossvalidation
based on the function `cv.glm()`. As we need to use the function `glm()`
for that, we have to use the manual calculations. But `cv.glm` expects a 
`data.frame` as input, so we need to adjust our previous code a bit.

The `cv.glm()` function has the following arguments: `data, glmfit, cost` and `K`. `K` is the number of CV folds: `K=sample size` is LOOCV. The
argument `data` is the dataset and `cost` is the error function: 
for example MSE, AUC and the misclassification error.

The output of `cv.glm()` contains an element `delta` with 2 values. The
first one is the raw extra-sample error, the second one is one corrected
for bias. We ignore that bias and use the raw values.

```{r use K fold crossvalidation}
trainX.svd <- svd(trainX)
U <- trainX.svd$u
D <- diag(trainX.svd$d)  # the loadings matrix
trainZ <- U%*%D 

# Avoid too long runtime
nvals <- seq(5,30)
cv.glm.pcr <- rep(NA, length(nvals))

for(i in seq_along(nvals)){
  # fitdata
  fitdata <- data.frame(trainY, trainZ[,1:nvals[i]])
  # make the model
  mod <- glm(trainY ~ ., data = fitdata)
  # do K fold crossvalidation. 
  cv.glm.pcr[i] <- cv.glm(fitdata, mod, cost = MSE, K = 20)$delta[1]
}

nPC.at.min <- nvals[which.min(cv.glm.pcr)]
plot(nvals, cv.glm.pcr, type = "l",
     ylab = "MSE", xlab = "number PCs")
abline(v = nPC.at.min, col = "red")
```

This outcome shows us that we can use approximately 12 PCs.

Now let’s validate this with the test dataset and see how well we are
doing with LOOCV. Since SVD is also an estimation procedure we will be
cheating if we perform SVD on the test dataset. We will use the loadings
of the training to construct PCs on the test: we are now sure we are
only using the validation set for testing.

```{r test on the test dataset}
# Create scores for the test using loadings of the validation
V <- trainX.svd$v
testZ <- testX %*% V

fitdata <- data.frame(trainY, trainZ[,1:12])
mod <- glm(trainY ~ ., data = fitdata)

preddata <- data.frame(testZ[, 1:12])

preds <- predict(mod, newdata = preddata)
MSE(testY, preds)


```

# Exercise: evaluation prediction models

#### 1. Dichotomize the training outcome variable, `trainY` at its 50% quantile. {-}
    
```{r}
trainY_bin <- ifelse( trainY > median(trainY), 1, 0)
table(trainY_bin)
testY_bin <- ifelse( testY > median(trainY), 1, 0)
table(testY_bin)
```

#### 2. Construct a prediction model using PC regression on the training dataset and select the optimum number of PCs using 5 folds cross validation with the misclassification (at cutoff $c=0.5$ ) error and the AUC as evaluation measures. Use the binary `trainY` as your response variable. {-}
    
```{r, message=FALSE}
K <- 5 # set K 
set.seed(1) # Use this if you want the exact same results
nPC <- 40 # max number of PCs, choose as you want with the 
          # maximum given by dim(trainZ)[2]
cv.pcr.error.bin = rep(NA,nPC)#We store our errors here

for (i in seq_len(nPC)) {
	fitdata <- data.frame(trainY_bin = trainY_bin,
	                      trainZ[,1:i])
	# Fit the model
	cv.pcr.mod1 <- suppressWarnings(
	  glm(trainY_bin ~ ., data = fitdata, 
	                   family="binomial")
	)
	# do K fold crossvalidation
	cv.pcr.error.bin[i] <- suppressWarnings(
	  cv.glm(fitdata,cv.pcr.mod1, 
	         cost = auc,K=K)$delta[1]
	  )
	# You use auc(), which is a function of the pROC package
	# It's possible that the code above produces a lot of messages
	# You can specify `message = FALSE` in the code chunk options to suppress
	# them (if you're working in RMarkdown)
}
```

Note that you don't have that much cases, and a ton of variables. This
can lead to lack of convergence in the internal fitting of the `glm()`.
It's one of the reasons we need to reduce dimensions! The function
`suppressWarnings()` suppresses all the warnings generated by ill fits.

You want to maximize the AUC, so you need to look for the maximum
value:
```{r}
nPC_at_max_AUC_CV <- c(1:nPC)[which.max(cv.pcr.error.bin)]
plot(cv.pcr.error.bin ,xlab = "n PCs",ylab = "AUC",type = 'l')
abline(v=nPC_at_max_AUC_CV,col="red")
```

Now we know the number of PCs, we can construct the model and 
test it against our test data.

```{r}
fitdata <- data.frame(trainY_bin = trainY_bin,
	                      trainZ[,1:nPC_at_max_AUC_CV])

finalmodel <- glm(trainY_bin ~ . ,
                  data = fitdata,family="binomial")

# Perform testing on the data
testdata <- data.frame(testZ[,1:nPC_at_max_AUC_CV])

# We predict the response (type = "response")
predPCR <- predict(finalmodel,
                   newdata=testdata,type = "response")

#We can now calculate the AUC 
#on the test data set just to see
#how good are we doing on test?
auc(testY_bin, predPCR)
```

#### 3. With your dichotomized outcome `trainY` construct a Lasso and ridge regression prediction model on the training dataset, select the optimum $\gamma$ by using 5 folds CV and the AUC as your evaluation measure. {-}

*Hint*: use `cv.glmnet()` function, for 5 folds CV, set `nfolds=5` and for the AUC set `type.measure="auc"`.
Go to `?cv.glmnet` for details.
    
```{r}
set.seed(5)
lasso_mod <- cv.glmnet(trainX, trainY_bin,
                       family="binomial",
                       type.measure="auc",nfolds=K,
                       alpha=1)
plot(lasso_mod)
```

We can look for the gamma values that give the best result. Here you have two possibilities :

1. look at `lambda.min`, which is the value of 
lambda that gives the best result for the crossvalidation.
2. look at `lambda.1se`, which is the largest value of
lambda such that the error is within 1 standard error
of the best result for the crossvalidation.
```{r}
#best gammas 
bestgamma.1se <- lasso_mod$lambda.1se
bestgamma.min <- lasso_mod$lambda.min
bestgamma.1se
bestgamma.min
```

Let's choose the best gamma for the prediction. You can
do this by passing the lambda value as the `s` argument
of the `predict()` function:

```{r}
predLasso_prob <- predict(lasso_mod, newx= testX,
                          s=bestgamma.min, type="response")

#how good are we doing on test? 
auc(testY_bin, as.vector(predLasso_prob))


```

We can now repeat this for the ridge regression:

```{r}
set.seed(5)
ridge_mod <- cv.glmnet(trainX, trainY_bin,
                       family="binomial",
                       type.measure="auc",nfolds=K,
                       alpha=0)
plot(ridge_mod)
```

Let's do the model evaluation:

```{r}
bestgamma.min <- ridge_mod$lambda.min

predRidge_prob <- predict(ridge_mod, newx= testX,
                          s=bestgamma.min, type="response")

#how good are we doing on test? 
auc(testY_bin, as.vector(predRidge_prob))
```

#### 4. For all the above prediction models, select the best model in terms of their estimated AUC (extra-sample error) on the training dataset. Confirm your choice of model on the test dataset. Look at the function `auc` of the package `pROC`. {-}
    
For me it looks like the lasso model performs the best on the test
data.

#### 5. Find the optimal $c$ of your best model in terms of misclassification error. {-}

After deciding on your choice of model based on its AUC, to actually use your model for prediction, you need an optimal cut-off point say $c\in [0,1]$, where when a predicted probability $p$ is greater than $c$, predict as high gene expression and when $p\leq c$ predict as low.
The optimal here means, $c$ has minimum misclassification error. 

For this, you need the function `MISERR` we defined above. You can 
again loop over a number of possible $c$ values, and check whether 
we get at a correct representation.

We chose the lasso as best, so:

```{r}
cvals <- seq(0.01,0.99,by = 0.01)
n <- length(cvals)

c_miss <- rep(NA, n)

for(i in seq_along(cvals)){
  c_miss[i] <- MISERR(testY_bin, 
                      predLasso_prob,
                      cvals[i])
}

best_c <- cvals[which.min(c_miss)]

plot(cvals, c_miss, type = "l")
abline(v = best_c, col = "red")

```

This shows a cutoff value of 0.47 would give the least amount of
misclassification error.
