---
title: "Analysis of High Dimensional Data - Lab 3"
subtitle: "Penalized regression techniques for high-dimensional data"
author: "Adapted by Milan Malfait"
date: "05 Nov 2020"
---

```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  out.width = "100%"
)
options(
  warnPartialMatchDollar = FALSE,
  warnPartialMatchAttr = FALSE,
  warnPartialMatchArgs = FALSE
)
```

***

```{r libraries, warning=FALSE, message=FALSE}
## install packages with:
## install.packages(c("glmnet", "pls", "NormalBetaPrime", "boot"))
library(NormalBetaPrime)
library(glmnet)
library(pls)
library(boot)
```


# Introduction

**In this lab session we will look at the following topics**

  - Demonstrate why low dimensional prediction modeling fails in high dimension.
  - Carry out Principal Component Regression (PCR)
  - Use `glmnet()` to carry out ridge regression, lasso and elastic net
  - Evaluation of these prediction models
  

## The dataset

In this practical, we will use the dataset `eyedata` provided by
the [__NormalBetaPrime__ package](https://cran.r-project.org/web/packages/NormalBetaPrime/index.html).
This dataset contains gene expression data of 200
genes for 120 samples. The data originates from microarray experiments
of mammalian eye tissue samples.

The dataset consists of two objects:

  - `genes`: a $120 \times 200$ matrix with the expression levels of 200 genes
  (columns) for 120 samples (rows)
  - `trim32`: a vector with 120 expression levels of the TRIM32 gene.


```{r load-data}
data(eyedata)
ls() # List objects in current R session
```

The goal of this exercise is to predict the expression levels of
TRIM32 from the expression levels of the 200 genes measured in the
microarray experiment. For this, it makes sense to start by constructing
centered (and possibly scaled) data. We store this in two matrices
`X` and `Y`:

```{r prepare-data}
X <- scale(genes, center = TRUE, scale = TRUE) 
Y <- scale(trim32, center = TRUE)
```

Remember that scaling avoids that differences in levels of magnitude
will give one variable (gene) more influence in the result. This has
been illustrated in the [second practical session](./Lab2-PCA.html) as well.
For the `Y` vector, this is less of an issue as we're talking about a single variable.
Not scaling will make the predictions interpretable as "deviations from the
mean".

## The curse of singularity

We begin by assuming that the predictors and the outcome have been
centered so that the intercept is 0.
We are presented with the usual regression model:

$$
Y_i=\beta_i X_{i1}+\dots+\beta_pX_{ip}+\epsilon_i \\ 
\text{ Or } \mathbf{Y}={\mathbf{X}}{\boldsymbol{\beta}} +{\boldsymbol{\epsilon}}
$$

Our goal is to get the least squares estimator of
${\boldsymbol{\beta}}$, given by

$$
\hat{{\boldsymbol{\beta}}}= (\mathbf{X}^T{\mathbf{X}})^{-1}{\mathbf{X}}^T{\mathbf{Y}}
$$

in which the $p \times p$ matrix
$({\mathbf{X}}^T{\mathbf{X}})^{-1}$ is crucial!
To be able to calculate the inverse of ${\mathbf{X}}^T \mathbf{X}$,
it has to be of full rank $p$, which would be 200 in this case.
Let's check this:

```{r singularity-problem, error=TRUE}
dim(X) # 120 x 200, so p > n!
qr(X)$rank

XtX <- crossprod(X) # calculates t(X) %*% X more efficiently
qr(XtX)$rank

# Try to invert using solve: 
solve(XtX)
```

We realize we cannot compute
$({\mathbf{X}}^T{\mathbf{X}})^{-1}$ because the rank of
$({\mathbf{X}}^T{\mathbf{X}})$ is less than $p$ hence we can’t
get $\hat{{\boldsymbol{\beta}}}$ by means of least squares! 
This is generally referred to as the __[singularity](https://www.statistics.com/glossary/singularity/) problem__.


# Principal component regression

A first way to deal with this singularity, is to bypass it using principal components.
Since $\min(n,p) = n = 120$, 
PCA will give `r min(dim(X))` components, each being a linear combination of the 
$p$ = `r ncol(X)` variables.
These `r min(dim(X))` PCs contain all information present in the original data.
We could as well use an approximation of ${\mathbf{X}}$, i.e using just a few ($k<120$) PCs.
So we use PCA as a method for reducing the dimensions while retaining
as much variation between the observations as possible.
Once we have these PCs, we can use them as variables in a linear regression model.

## Classic linear regression on PCs

We first compute the PCA on the data with `prcomp`.
We will use an arbitrary cutoff of $k = 4$ PCs to illustrate the process of performing regression on the PCs.

```{r PC-regression}
k <- 4 # Arbitrarily chosen k=4
pca <- prcomp(X)
Vk <- pca$rotation[, 1:k] # the loadings matrix
Zk <- pca$x[, 1:k] # the scores matrix

# Use the scores in classic linear regression
pcr_model1 <- lm(Y ~ Zk)
summary(pcr_model1)
```

As $\mathbf{X}$ and $\mathbf{Y}$ are centered, the intercept is 
approximately 0.

The output shows that PC1 and PC4 have a $\beta$ estimate that 
differs significantly from 0 (at $p < 0.05$), but the results can't be readily 
interpreted, since we have no immediate interpretation of the PCs.


## Using the package `pls`

PCR can also be performed using the `pcr()` function from the
package *[pls](https://CRAN.R-project.org/package=pls)*
__directly on the data__ (so without having to first perform the PCA manually).
When using this function, you have to keep a few things in mind:

  1. the number of components (PCs) to use is passed with the argument `ncomp`
  2. the function allows you to scale (set `scale = TRUE`) and
  center (set `center = TRUE`) the predictors first (in the example here, $\mathbf{X}$ has already been centered and scaled).
  
You can use the function `pcr()` in much the same way as you would
use `lm()`. The resulting fit can easily be examined using the 
function `summary()`, but the output looks quite different from
what you would get from `lm`.

```{r PC-regression-pls-package}
# X is already scaled and centered, so that's not needed.
pcr_model2 <- pcr(Y ~ X, ncomp = 4)
summary(pcr_model2)
```

First of all the output shows you the data dimensions and the fitting
method used. In this case, that is PC calculation based on SVD. The
`summary()` function also provides the percentage of variance
explained in the predictors and in the response using different numbers
of components. For example, the first PC only captures 61.22% of all
the variance, or information in the predictors and it explains 62.9%
of the variance in the outcome. Note that for both methods the choice of
the number of principal components was arbitrary chosen to be 4.

At a later stage, we will look at how to choose the number of components
that has the __smallest prediction error__.


# Ridges, Lassos and Elastic Nets {#elnet-theory}

Ridge regression, lasso regression and elastic nets are all closely
related techniques, based on the same idea: add a penalty term to 
the estimating function so $({\mathbf{X}}^T{\mathbf{X}})$
becomes full rank again and is invertible. Two different penalty 
terms or regularization methods can be used:

1. L1 regularization: this regularization adds a term ${\gamma_1\|\boldsymbol{\beta}\|_{1}}$ to the estimating equation.
The term will add a penalty based on the *absolute value* of the
magnitude of the coefficients. This is used by the __lasso regression__
 
$$
 \hat{\boldsymbol{\beta}}^{\text{lasso}} = \text{argmin}_{\boldsymbol{\beta}}\displaystyle({(\mathbf{Y}-\mathbf{X}\boldsymbol{\beta})^T(\mathbf{Y}-\mathbf{X}\boldsymbol{\beta})+{\gamma_1\|\boldsymbol{\beta}\|_{1}}}\displaystyle)
$$

2. L2 regularization: this regularization adds a term ${\gamma_2\|\boldsymbol{\beta}\|_{2}^{2}}$ to the estimating equation.
The penalty term is based on the square of the magnitude of the 
coefficients. This is used by __ridge regression__.

$$
 \hat{\boldsymbol{\beta}}^{\text{ridge}} = \text{argmin}_{\boldsymbol{\beta}}\displaystyle({(\mathbf{Y}-\mathbf{X}\boldsymbol{\beta})^T(\mathbf{Y}-\mathbf{X}\boldsymbol{\beta})+{\gamma_2\|\boldsymbol{\beta}\|_{2}^{2}}}\displaystyle)
$$

Elastic nets combine both types of regularizations. It does so by 
introducing a $\alpha$ mixing parameter that essentially combines
the L1 and L2 norms in a weighted average.

$$
 \hat{\boldsymbol{\beta}}^{\text{el.net}} = \text{argmin}_{\boldsymbol{\beta}}\displaystyle({(\mathbf{Y}-\mathbf{X}\boldsymbol{\beta})^{T}(\mathbf{Y}-\mathbf{X}\boldsymbol{\beta})+{\alpha \gamma_1\|\boldsymbol{\beta}\|_{1}}+ {(1 - \alpha)\gamma_2\|\boldsymbol{\beta}\|_{2}^{2}}}\displaystyle)
$$



# Exercise: Verification of ridge regression

In least square regression the minimization of the estimation function
$|{\mathbf{Y} - \mathbf{X} \boldsymbol{\beta}}\|^{2}_{2}$ leads to the solution ${\boldsymbol{\hat{\beta}}=(\mathbf{X^TX})^{-1}\mathbf{X^TY}}$. 

For the penalized least squares criterion used by ridge regression, you minimize 
$\|{\mathbf{Y}-\mathbf{X}\boldsymbol{\beta}\|^{2}_{2}}+\gamma{\boldsymbol{\|\beta\|^{2}_{2}}}$
which leads to following solution:

$$
{\boldsymbol{\hat{\beta}}=(\mathbf{X^TX}}+\gamma{\mathbf{I}})^{-1}{\mathbf{X^TY}}
$$

where $\mathbf{I}$ is the $p \times p$ identity matrix.

The ridge parameter $\gamma$ *shrinks* the coefficients towards 0, with $\gamma = 0$ being equivalent to OLS (no shrinkage) and $\gamma = +\infty$ being equivalent to setting all $\hat{\beta}$'s to 0.
The optimal parameter lies somewhere in between and needs to be tuned by the user.


## Tasks {-}

Solve the following exercises using R.

#### 1. Verify that ${\mathbf{(X^TX}}+\gamma{\mathbf{I}})$ has rank $200$, for any $\gamma>0$ of your choice. {-}

<details><summary>Solution</summary>
```{r}
XtX <- crossprod(X)
p <- ncol(X)
gamma <- 2 # My choice

# Compute penalized matrix
XtX_gammaI <- XtX + (gamma * diag(p))
dim(XtX_gammaI)
qr(XtX_gammaI)$rank == 200 # indeed
```
</details>


#### 2. Check that the inverse of ${\mathbf{(X^TX}}+\gamma{\mathbf{I}})$ can be computed. {-}

<details><summary>Solution</summary>
```{r}
# Yes, it can be computed (no error)
XtX_gammaI_inv <- solve(XtX_gammaI)
str(XtX_gammaI_inv)
```
</details>


#### 3. Finally, compute ${\boldsymbol{\hat{\beta}}=(\mathbf{X^TX}}+\gamma{\mathbf{I}})^{-1}{\mathbf{X^TY}}$. {-}

<details><summary>Solution</summary>
```{r ridge-beta-estimates}
## Calculate ridge beta estimates
## Use `drop` to drop dimensions and create vector
ridge_betas <- drop(XtX_gammaI_inv %*% t(X) %*% Y)
length(ridge_betas) # one for every gene
summary(ridge_betas)
```

We have now manually calculated the ridge regression estimates.

</details>


<!-- If we perform SVD on ${\mathbf{X}}$ we get -->
<!-- ${\mathbf{X}}={\mathbf{UDV^T}}={\mathbf{ZV^T}}$. -->
<!-- Replacing ${\mathbf{X}}$ with ${\mathbf{ZV^T}}$ gives -->
<!-- ${\boldsymbol{\hat{\beta}}=\mathbf{V(Z^TZ}}+\gamma{\mathbf{I}})^{-1}{\mathbf{Z^TY}}$ -->

<!-- #### 4. Check to see that ${\mathbf{(X^TX}}+\gamma{\mathbf{I}})$ has a higher dimension than ${\mathbf{(Z^TZ}}+\gamma{\mathbf{I}})$. {-} -->

<!-- And therefore inverting ${\mathbf{(Z^TZ}}+\gamma{\mathbf{I}})$ is much faster than inverting -->
<!-- ${\mathbf{(X^TX}}+\gamma{\mathbf{I}})$ (You don’t have to check this). -->

<!-- <details><summary>Solution</summary> -->
<!-- ```{r} -->
<!-- X.svd<-  svd(X) -->
<!-- Uk <- X.svd$u -->
<!-- Dk <- diag(X.svd$d) -->
<!-- Zk <- Uk%*%Dk -->
<!-- Vk <- X.svd$v -->
<!-- p<-dim(Zk)[1] -->
<!-- tZZ_gammaI<-(t(Zk)%*%Zk)+gamma*diag(p) -->
<!-- dim(tZZ_gammaI) -->

<!-- # Calculate the beta -->
<!-- hatbeta_svd <- Vk %*% solve(tZZ_gammaI) %*% t(Zk) %*% Y -->

<!-- # Check whether the same -->
<!-- comparison <- cbind(hatbeta, hatbeta_svd) -->
<!-- head(comparison) # to not show everything  -->
<!-- ``` -->
<!-- </details> -->

<!-- We therefore conclude that -->
<!-- ${\boldsymbol{\hat{\beta}}=\mathbf{V(Z^TZ}}+\gamma{\mathbf{I}})^{-1}{\mathbf{Z^TY}}$ -->
<!-- is a faster way for computing ${\boldsymbol{\hat{\beta}}}$. -->



# Performing ridge and lasso regression with `glmnet`

The package *[glmnet](https://CRAN.R-project.org/package=glmnet)* provides a
function `glmnet()` that allows you to fit all three types of regressions. Which
type is used, can be determined by specifying the `alpha` argument. For a
__ridge regression__, you set `alpha` to 0, and for a __lasso regression__ you
set `alpha` to 1. Other `alpha` values between 0 and 1 will fit a form of
elastic net. This function has slightly different syntax from the other
model-fitting functions. To be able to use it, you have to pass a `x` matrix as
well as a `y` vector, and you don't use the formula syntax.

The gamma value, which controls the "strength" of the penalty, can be passed by
the argument `lambda` (notation isn't always consistent between text books and
software...). The function `glmnet()` can also carry out a search for finding
the best gamma value for a fit. This can be done by passing multiple values to
the argument `lambda`. If not supplied, `glmnet` will generate a range of values
itself, based on the data whereby the number of values can be controlled with
the `nlambda` argument. This is generally the recommended way to use `glmnet`,
see `?glmnet` for details.

For a thorough introduction to the __glmnet__ package and elastic net models in
general, see the
[glmnet introduction vignette](https://cran.r-project.org/web/packages/glmnet/vignettes/glmnet.pdf)


## Demonstration: Ridge regression {-}

Let's perform a ridge regression in order to predict expression levels
of the TRIM32 gene using the 200 gene probes data. We can start by
using a $\gamma$ value of 2.

```{r glmnet-ridge-regression}
gamma <- 2
ridge_model <- glmnet(X, Y, alpha = 0, lambda = gamma)

# have a look at the first 10 coefficients
coef(ridge_model)[1:10]
```

The first coefficient is the intercept, and is again essentially 0. But
a value of 2 for $\gamma$ might not be the best choice, so let's see how
the coefficients change with different values for $\gamma$.

We will create a *grid* of $\gamma$ values, i.e. a range of values that will be
used as input for the `glmnet` function. Note that this function can take a
vector of values as input for the `lambda` argument, allowing to fit multiple
models with the same input data but different hyperparameters.

```{r ridge-regression-grid-search}
grid <- seq(1, 1000, by = 10)  # 1 to 1000 with steps of 10
ridge_mod_grid <- glmnet(X, Y, alpha = 0, lambda = grid)

# Plot the coefficients against the (natural) LOG lambda sequence!
# see ?plot.glmnet
plot(ridge_mod_grid, xvar = "lambda", xlab = "log(gamma)")
# add a vertical line at gamma = 2
text(log(gamma), -0.05, labels = expression(gamma == 2), 
     adj = -0.5, col = "firebrick")
abline(v = log(gamma), col = "firebrick", lwd = 2)
```

This plot is known as a __coefficient profile plot__, each colored line
represents a coefficient $\hat{\beta}$ from the regression model and shows how
they change with increased values of $\gamma$ (on the log-scale)
^[Note: `log()` in R is the __natural logarithm__ by default (base $e$) and we
will also use this notation in the text (like the x-axis title on the plot above).
This might be different from the notation that you're used to ($\ln()$).
To take logarithms with a different base in R you can specify the `base = `
argument of `log` or use the shorthand functions `log10(x)` and `log2(x)` for
base 10 and 2, respectively].

Note that for higher values $\gamma$, the coefficient estimates become closer to 0,
showing the *shrinkage* effect of the ridge penalty.

Similar to the PC regression example, we chose $\gamma=2$ and the grid rather
arbitrarily. We will see subsequently, how to choose $\gamma$ that minimizes the
prediction error.


# Exercise: Lasso regression

Lasso regression is also a form of penalized regression, but we do not have an
analytic solution of $\hat{{\boldsymbol{\beta}}}$ as in least squares
and ridge regression. In order to fit a lasso model, we once again use
the `glmnet()` function. However, this time we use the argument
`alpha = 1`


## Tasks {-}

#### 1. Verify that setting `alpha = 1` indeed corresponds to lasso regression using the equations from [Section 3](#elnet-theory). {-}


#### 2. Perform a lasso regression with the `glmnet` function with `Y` the response and `X` the predictors. {-}

You do not have to provide a custom sequence of $\gamma$ (`lambda`) values here
but can instead rely on `glmnet`'s default behaviour of choosing the grid of
$\gamma$ values based on the data (see `?glmnet` for more details).

<details><summary>Solution</summary>
```{r glmnet-lasso-regression}
# Note that the glmnet() function can supply gamma automatically
# By default it uses a sequence of 100 lambda values
lasso_model <- glmnet(X, Y, alpha = 1)
```
</details>


#### 3. Make the coefficient profile plot and interpret. {-}

<details><summary>Solution</summary>

```{r}
plot(lasso_model, xvar = "lambda", xlab = "log(gamma)")
```

Note that the number of non-zero coefficients is indicated at the top of the plot.
In the case of lasso-regression the regularization is much less smooth compared
to the ridge regression, with some coefficients increasing for higher $\gamma$
before sharply dropping to zero.
In contrast to ridge, lasso eventually shrinks all coefficients to 0.

</details>


# Evaluation of prediction models and tuning hyperparameters

First we will split our original data in a training and test set to validate our
model. The training set will be used to train the model and tune the
hyperparameters, while the test set will be used to evaluate the
__out-of-sample__ performance of our final model. If we would use the same data
to both fit and test the model, we would get biased results.

Before we begin, we use the `set.seed()` function in order to set a seed
for R’s random number generator, so that we will all obtain precisely
the same results as those shown below. It is generally good practice to
set a random seed when performing an analysis such as cross-validation
that contains an element of randomness, so that the results obtained can
be reproduced at a later time.

We begin by using the `sample()` function to split the set of samples into two
subsets, by selecting a random subset of 80 observations out of the original 120
observations. We refer to these observations as the __training__ set. The rest
of the observations will be used as the __test__ set.

```{r create-training-set}
set.seed(1)
# Sample 80 random IDs from the rows of X (120 total)
trainID <- sample(nrow(X), 80)

# Training data
trainX <- X[trainID, ]
trainY <- Y[trainID]

# Test data
testX <- X[-trainID, ]
testY <- Y[-trainID]
```

To make fitting the models a bit easier later, we will also create 2 data.frames
combining the response and predictors for the training and test data.

```{r}
train_data <- data.frame("TRIM32" = trainY, trainX)
test_data <- data.frame("TRIM32" = testY, testX)

## Glancing at the data structure: for the first 10 columns only
str(train_data[, 1:10])
```


## Model evaluation

We are interested in the __out-of-sample__ error of our models,
i.e. how good our model does on unseen data.
__This will allow us to compare different *classes* of models__.
For continuous outcomes we will use the __mean squared error (MSE)__
(or its square-root version, the RMSE).

The evaluation will allow us to compare the performance of different types of
models, e.g. PC regression, ridge regression and lasso regression, on our data.
However, we still need to find the optimal model within each of these classes,
by selecting the best hyperparameter (number of PCs for PC regression and $\gamma$
for lasso and ridge).
For that we will use 
[*$k$-fold Cross Validation*](https://en.wikipedia.org/wiki/Cross-validation_(statistics))
on our training set.


## Tuning hyperparameters

The test set is only used to evaluate the *final* model.
To achieve this final model, we need to find the optimal hyperparameters,
i.e. the hyperparameters that best generalize the model to unseen data.
We can estimate this by using *k-fold cross validation* (CV$_k$) on
the training data.

The CV$_k$ estimates can be automatically computed for any
generalized linear model (generated with `glm()` and by extension `glmnet()`) 
using the `cv.glm()` function from the
*[boot](https://CRAN.R-project.org/package=boot)* package.


# Example: PC regression evaluation

We start with the PC regression and look for the optimal number of PCs that minimizes
the MSE using $k$-fold Cross validation.
We then use this optimal number of PCs to train the final model and evaluate it
on the test data.


## k-fold Cross Validation to tune number of components

Conveniently, the `pcr` function from the `pls` package has an implementation for
k-fold Cross Validation. We simply need to set `validation = CV` and `segments = 20`
to perform 20-fold Cross Validation with PC regression.
If we don't specify `ncomp`, `pcr` will select the maximum number of PCs that can
be used for the CV.

Note that our training data `trainX` consists of 80 observations (rows).
If we perform 20-fold CV, that means we will split the data in 20 groups, so
each group will consist of 4 observations. At each CV cycle, one group will be left
out and the model will be trained on the remaining groups. This leaves us with
76 training observations for each CV cycle, so the maximal number of components
that can be used in the linear regression is 75.

```{r pcr-kCV}
## Set seed for reproducibility, kCV is a random process!
set.seed(123)

K <- 20

## The 'Y ~ .' notation means: fit Y by every other variable in the data
pcr_cv <- pcr(TRIM32 ~ ., data = train_data, validation = "CV", segments = K)
summary(pcr_cv)
```

We can plot the *root mean squared error of prediction* (RMSEP) for each number
of components as follows.

```{r pcr_cv-plot}
plot(pcr_cv, plottype = "validation")
```

The `pls` package also has a function `selectNcomp` to select the optimal number of components.
Here we use the "one-sigma" method, which returns the lowest number of components
for which the RMSE is within one standard error of the absolute minimum.
The function also allows plotting the result by specifying `plot = TRUE`.

```{r pcr-optimal-ncomp}
optimal_ncomp <- selectNcomp(pcr_cv, method = "onesigma", plot = TRUE)
```

This outcome shows us that the optimal number of components for our model is
`r optimal_ncomp`.


## Validation on test data

We now use our optimal number of components to train the final PCR model.
This model is then validated on by generating predictions for the test data and
calculating the MSE.

We define a custom function to calculate the MSE.
Note that there is also an `MSEP` function in the `pls` package which does the
prediction and MSE calculation in one go.
But our own function will come in handy later for lasso and ridge regression.

```{r MSE}
# Mean Squared Error
## obs: observations; pred: predictions
MSE <- function(obs, pred){
  mean((drop(obs) - drop(pred))^2)
}
```

```{r final_pcr_model}
final_pcr_model <- pcr(TRIM32 ~ ., data = train_data, ncomp = optimal_ncomp)
pcr_preds <- predict(final_pcr_model, newdata = test_data, ncomp = optimal_ncomp)
(pcr_mse <- MSE(testY, pcr_preds))
```

For comparison, `MSEP` also gives the MSE for a model including only an intercept.
This value on its own does not tell us very much, but we can use it to compare our
PCR model with other types of models later.

Finally, we plot the predicted values for our response variable (the TRIM32 gene expression)
against the actual observed values from our test set.

```{r pcr-predplot}
predplot(final_pcr_model, newdata = test_data, line = TRUE)
```



# Exercise: evaluate and compare prediction models

#### 1. Perform a lasso regression with 20-fold Cross Validation on the training data (`trainX`, `trainY`). Plot the results and select the optimal `lambda` ($\gamma$) parameter. Fit a final model with the selected `lambda` and validate it on the test data. {-}

*Hint*: use the `cv.glmnet()` function, for 20 folds CV, set `nfolds = 20` and 
to use the MSE metric set `type.measure = "mse"`.
Go to `?cv.glmnet` for details.

<details><summary>Solution</summary>

```{r lasso-cv}
set.seed(123)
lasso_cv <- cv.glmnet(trainX, trainY, alpha = 1, 
                      nfolds = K, type.measure = "mse")
lasso_cv
plot(lasso_cv)
```

Note that we can extract the fitted lasso regression object from the CV result
and make the coefficient profile plot as before.

```{r lasso-cv-coefficient-profile}
plot(lasso_cv$glmnet.fit, xvar = "lambda")
```

We can look for the gamma values that give the best result. 
Here you have two possibilities :

1. `lambda.min`: the value of  $\gamma$ that gives the best result for the crossvalidation.
2. `lambda.1se`: the largest value of $\gamma$ such that the MSE is within 1 standard error
of the best result from the cross validation.

```{r}
lasso_cv$lambda.min
lasso_cv$lambda.1se
```

We will (rather arbitrarily) use `lambda.min` here to fit the final model and generate predictions on the test data.
Note that we don't actually have to redo the fitting, we can just use our existing
`lasso_cv` object, which already contains the fitted models for a range of `lambda` values.
We can use the `predict` function and specify the `s` argument (which confusingly sets `lambda` in this case)  to make predictions on the test data.

```{r}
lasso_preds <- predict(lasso_cv, s = lasso_cv$lambda.min, newx = testX)
## Calculate MSE
(lasso_mse <- MSE(testY, lasso_preds))
```
</details>


#### 2. Do the same for ridge regression. {-}

<details><summary>Solution</summary>

```{r ridge-cv}
set.seed(123)
ridge_cv <- cv.glmnet(trainX, trainY, alpha = 0, 
                      nfolds = K, type.measure = "mse")
ridge_cv
plot(ridge_cv)
```

Note that we can extract the fitted ridge regression object from the CV result
and make the coefficient profile plot as before.

```{r ridge-cv-coefficient-profile}
plot(ridge_cv$glmnet.fit, xvar = "lambda")
```

We can look for the gamma values that give the best result. 
Here you have two possibilities :

1. `lambda.min`: the value of  $\gamma$ that gives the best result for the crossvalidation.
2. `lambda.1se`: the largest value of $\gamma$ such that the MSE is within 1 standard error
of the best result from the cross validation.

```{r}
ridge_cv$lambda.min
ridge_cv$lambda.1se
```

We will (rather arbitrarily) use `lambda.min` here to fit the final model and generate predictions on the test data.
Note that we don't actually have to redo the fitting, we can just use our existing
`ridge_cv` object, which already contains the fitted models for a range of `lambda` values.
We can use the `predict` function and specify the `s` argument (which confusingly sets `lambda` in this case)  to make predictions on the test data.

```{r ridge-predictions}
ridge_preds <- predict(ridge_cv, s = ridge_cv$lambda.min, newx = testX)
## Calculate MSE
(ridge_mse <- MSE(testY, ridge_preds))
```

</details>


#### 3. Which of the models considered (PCR, lasso, ridge) performs best?. {-}
    
<details><summary>Solution</summary>

Based on the MSE, the ridge model performs best on the test data.

```{r, echo=FALSE}
knitr::kable(
  data.frame(
    "Model" = c("PCR", "Lasso", "Ridge"),
    "MSE" = c(pcr_mse, lasso_mse, ridge_mse)
  )
)
```
</details>
