---
title: "Analysis of High Dimensional Data - Lab 4"
subtitle: "Sparse PCA and LDA"
author: "Adapted by Milan Malfait"
date: "19 Nov 2020"
references:
- id: alon1999broad
  type: article-journal
  author:
  - family: Alon
    given: Uri
  - family: Barkai
    given: Naama
  - family: Notterman
    given: Daniel A
  - family: Gish
    given: Kurt
  - family: Ybarra
    given: Suzanne
  - family: Mack
    given: Daniel
  - family: Levine
    given: Arnold J
  issued:
  - year: 1999
  title: Broad patterns of gene expression revealed by clustering analysis of tumor
    and normal colon tissues probed by oligonucleotide arrays
  container-title: Proceedings of the National Academy of Sciences
  publisher: National Acad Sciences
  page: 6745-6750
  volume: '96'
  issue: '12'
---

```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 8,
  fig.asp = 0.618,
  out.width = "100%"
)
```

***

```{r libraries, warning=FALSE, message=FALSE}
## install packages with:
## install.packages(c("glmnet", "MASS"))
library(glmnet)
library(MASS)
```


# Introduction

**In this lab session we will look at the following topics**

  - Methods to set some of the loadings exactly to zero in a PCA.
  - Use `glmnet()` to add penalties on principal component loadings.
  - Use LDA to understand differences between groups in a high dimensional space.
  
## The dataset {-}

In this practical session, we use the dataset by @alon1999broad on
gene expression levels in 40 tumour and 22 normal colon tissue samples.
They checked a total of 6500 human genes using the Affymetrix oligonucleotide array.
You can read the data in as follows:

```{r load-data}
Alon1999 <- read.csv("https://github.com/statOmics/HDA2020/raw/data/Alon1999.csv")
str(Alon1999[, 1:10])
table(Alon1999$Y)
```

The dataset contains one variable named `Y` with the values `t` and `n`.
This variable indicates whether the sample came from tumourous (`t`) or
normal (`n`) tissue.

The goal of this practical is to find the best subset/combination of genes to detect tumourous tissue.
As in @alon1999broad, we use the 2000 genes with the highest minimal intensity
across the samples.

# Sparse PCA

In order to work easily with the data, first construct a scaled matrix
`X` and a vector `Y` which gives the scaled predictors and the response
variable:

```{r}
X <- scale(Alon1999[, -1])
Y <- as.factor(Alon1999[, 1])
```

Use these objects to solve the following exercises.

## Exercises {-}

#### 1. Perform a SVD on `X`, and store the scores of the PCs in a matrix `Z`. {-}

<details><summary>Solution</summary>

Using `svd`:

```{r}
svd_X <- svd(X)
Z <- svd_X$u %*% diag(svd_X$d) # Calculate the scores
V <- svd_X$v                 # Calculate the loadings
```

Using `prcomp`:

```{r}
## X is already centered and scaled so no need to do again
pca_x <- prcomp(X, center = FALSE, scale. = FALSE)
## The scores are given by `pca_x$x`, the loadings are `pca_x$rotation`
```

</details>


#### 2. Plot the singular values and confirm that the first and second PCs can approximate the data to some extent. {-}

<details><summary>Solution</summary>

```{r pca-singular_values-plot}
## Plotting parameters
par(pch = 19, mfrow = c(1, 2))

plot(svd_X$d, type = "b", ylab = "Singular values", xlab = "PCs")

## Percentage variance explained for each PC
var_explained <- svd_X$d^2 / sum(svd_X$d^2)
plot(var_explained,
  type = "b", ylab = "Percent variance explained", xlab = "PCs",
  col = 2
)
```

</details>


#### 3. Plot the first two PCs and use different colours for tumor/normal tissue. {-}

In order to plot different colors and add a legend with base `R` plotting, you can do the following:

```{r}
par(mfrow = c(1, 1))
cols <- c("n" = "red", "t" = "blue")
plot(X[, 1], X[, 2], col = cols[Y], pch = 19)
legend("topleft", c("Normal", "Tumor"),
  col = c("red", "blue"),
  pch = 19, title = "Tissue"
)
```

This plots the first two dimensions of the `X` (!) matrix with solid
points (`pch = 19`), and the color red for normal tissue and blue for
tumorous tissue. You can adapt this code to create the proper plot.

<details><summary>Solution</summary>

```{r pca-plot}
cols <- c("n" = "red", "t" = "blue")
plot(Z[, 1], Z[, 2],
  col = cols[Y],
  xlab = "PC1", ylab = "PC2", pch = 19
)
legend("topleft", c("Normal", "Tumor"),
  col = c("red", "blue"),
  pch = 19, title = "Tissue"
)
```

__Interpretation:__ using only the first 2 PCs does not seem to separate the tumour and normal cases clearly.

</details>


#### 4. Plot histograms of the loadings of the first and second PCs. Which loadings are the most important? {-}

You can use the `hist` function to plot a histogram.
Be sure to you use an appropriate value for the `breaks` argument.

<details><summary>Solution</summary>

```{r pc-loadings, fig.asp = 1.2}
par(mfrow = c(2, 1))
# First
hist(V[, 1], breaks = 50, xlab = "PC 1 loadings", main = "")
# Add vertical line at 95% quantile
abline(v = quantile(V[, 1], 0.95), col = "red", lwd = 2)

# Second
hist(V[, 2], breaks = 50, xlab = "PC 2 loadings", main = "")
abline(v = c(
  quantile(V[, 2], 0.05),
  quantile(V[, 2], 0.95)
), col = "red", lwd = 2)
```

Vertical lines were added at the 95th percentile for PC1 and the 5th and 95th percentiles for PC2 to reflect where the "highest" (in absolute value) loadings are situated (no negative loadings for PC1, so only showing the 95th percentile).

__Interpretation:__ remember that the PC loadings reflect the *contributions* of each feature (in this case: gene) to the PC.
From these histograms it should be clear that only a minor fraction of the genes are really driving these first 2 PCs, especially for PC 2 (where the bulk of genes has loadings close to 0).

</details>


#### 5. We know that the first PC $\mathbf{Z_1}$, is given by {-}

  $$
  \mathbf{Z_1}=\mathbf{X} \mathbf{V_1}
  $$

  Where $\mathbf{V_1}$ are the loadings of the first PC. If we put this in regression notation, we get

  $$
  \mathbf{Y}=\mathbf{X}\boldsymbol{\beta}
  $$ 

  where $\boldsymbol{\beta}$ now represent the $\mathbf{V_1}$ loadings, and
  $\mathbf{Y}$ is $\mathbf{Z_1}$.

  Recall that the ridge regression solution for $\boldsymbol{\beta}$ is given by

  $$
  \boldsymbol{\beta}_{\text{ridge}}
    = (\mathbf{X^TX}+\gamma\mathbf{I})^{-1}\mathbf{X}^T\mathbf Y
  $$

  __Question:__ Replace $\mathbf{Y}$ with $\mathbf{Z_1}$ and verify in 
  `R` that

  $$
  \mathbf V_1 = 
    \frac{\boldsymbol\beta_{\text{ridge}}}{\|\boldsymbol\beta_{\text{ridge}}\|_2}
  $$

  for any $\gamma > 0$ of your choice.
  Remember that
  $\|\boldsymbol\beta_{\text{ridge}}\|_2 = \sqrt{\sum_{j=1}^p \beta_j^2}$

<details><summary>Solution</summary>

```{r, cache=TRUE}
p <- dim(X)[2]

# Let's take a ridiculously large gamma: 200
tXX_gamma_I <- t(X) %*% X + 200 * diag(p)

## This might take a while to calculate
beta_ridge <- solve(tXX_gamma_I) %*% t(X) %*% Z[, 1]

#||beta_ridge||_2
mag_beta_ridge <- sqrt(sum(beta_ridge^2))
```

For comparison, let's plot
$\boldsymbol\beta_{\text{ridge}} / \|\boldsymbol\beta_{\text{ridge}}\|_2$
against the PC1 loadings $\mathbf V_1$.

```{r beta_ridge-vs-V1-plot}
par(mfrow = c(1, 1))

# Plot against the loadings
plot(svd_X$v[, 1], beta_ridge / mag_beta_ridge,
  xlab = expression("V"[1]),
  ylab = expression(beta["ridge"] / paste("||", beta, "||")[2]),
  pch = 19
)
# Or just simply take the difference between them
max(abs(svd_X$v[, 1] - beta_ridge / mag_beta_ridge))
```

</details>

  Then you've proven that the loadings of the PCs can be computed from the
  ridge regression coefficients.
  
  We can now move on to sparse PCA, where we use penalised regression to set some of the loadings ($\boldsymbol{\beta}$s) to zero.

#### 6. We have seen elastic net type penalties. If we call the loadings of the first PC as $\boldsymbol{\beta}$ and denote PC1 as $\mathbf{Y}$, we saw that $\boldsymbol{\beta}$ can be derived by minimising the SSE: {-}

  $$
  \text{SSE}=\|\mathbf Y-\mathbf{X}\boldsymbol \beta\|^2_2+\gamma\|\boldsymbol \beta\|^2_2 \text{ for any } \gamma>0).
  $$

  Note that this equality holds for any positive $\gamma$. So we can't penalise the $\boldsymbol\beta$s not being zero by choosing a different $\gamma$.
  Remember that for ridge regression, the $\beta$'s only become 0 for $\gamma = \infty$.
  Fortunately we have other tools. 

  In addition to the $L_2$ penalization, we can use the $L_1$ penalization of Lasso. This allows us to force some of the $\boldsymbol\beta$s to become zero. The new SSE will be of the form:

  $$
  \text{SSE}=\|\mathbf Y-\mathbf{X}\boldsymbol \beta\|^2_2+\gamma\|\boldsymbol \beta\|^2_2 +\gamma_1\|\boldsymbol \beta\|_1.
  $$

  This is exactly the elastic net SSE, and $\gamma_1$ is the Lasso type penalty that sets loadings to zero. 

  Now use the `glmnet` and `cv.glmnet` functions to select an appropriate number of non-zero loadings for the first and second PCs.
  Use `alpha = 0.5` in your elastic net models and use `Z1` and `Z2` as the response variables (you should fit 2 separate models).

<details><summary>Solution</summary>

```{r PC-cv_glmnet}
par(mfrow = c(1, 2))
# For PC1
set.seed(45)
fit_loadings1 <- cv.glmnet(X, Z[, 1],
  alpha = 0.5, nfolds = 5
)
plot(fit_loadings1, main = "PC1")

# For PC2
set.seed(45)
fit_loadings2 <- cv.glmnet(X, Z[, 2], alpha = 0.5, nfolds = 5)
plot(fit_loadings2, main = "PC2")
```

To see how many features are important for each fit, we can make coefficient profile plots.
Note that the actual `glmnet` fit objects are included in the `cv.glmnet` objects under `$glmnet.fit`.

I added vertical dashed lines at `lambda.min` and `lambda.1se`.

```{r PC-glmnet-coefficient-plots, fig.asp = 1.2}
par(mfrow = c(2, 1))
plot(fit_loadings1$glmnet.fit, main = "PC1", xvar = "lambda")
abline(v = log(fit_loadings1$lambda.min), lty = 3)
abline(v = log(fit_loadings1$lambda.1se), lty = 3)
plot(fit_loadings2$glmnet.fit, main = "PC2", xvar = "lambda")
abline(v = log(fit_loadings2$lambda.min), lty = 3)
abline(v = log(fit_loadings2$lambda.1se), lty = 3)
```

To get the exact number of non-zero coefficients for `lambda.min` and `lambda.1se`, just print the `cv.glmnet` objects.
```{r}
fit_loadings1
fit_loadings2
```

__Interpretation:__ for PC1, we see that around 90 to 100 genes are most important, based on the range of $\gamma$ (`lambda`) values between `lambda.min` and `lambda.1se`.
Similarly, for PC2 we get around 65 - 80 genes.
With this information, we can now choose one of the `lambda` values and construct PCs that will have non-zero loadings for only a few genes.

</details>


#### 7. Plot your newly derived first and second PCs and use different colors for the tumor and normal tissues. How well do these new PCs separate the response classes? Compare this to the plot in exercise 3. Formulate a conclusion based on the two graphs. {-}

Use `lambda.1se` as your choice for $\gamma$.
You can extract the coefficients ($\beta$'s) from the `cv.glmnet` objects using the `coef` function, set the `s` argument to the chosen $\gamma$.
This will return a *sparse matrix* by default, so you might want to use `as.vector` to convert to a more friendly format.

<details><summary>Solution</summary>

```{r sparse-PCA-plots, fig.width = 9}
sparse_loadings1 <- as.vector(coef(fit_loadings1, s = fit_loadings1$lambda.1se))
sparse_loadings2 <- as.vector(coef(fit_loadings2, s = fit_loadings2$lambda.1se))

## How many non-zero loadings do we have (excluding the intercept)?
(non_zero1 <- sum(abs(sparse_loadings1[-1]) > 0))
(non_zero2 <- sum(abs(sparse_loadings2[-1]) > 0))

SPC1 <- X %*% sparse_loadings1[-1] # without the intercept
SPC2 <- X %*% sparse_loadings2[-1] # without the intercept

par(mfrow = c(1, 2))
plot(Z[, 1], Z[, 2],
  col = cols[Y], xlab = "PC1", ylab = "PC2", pch = 16,
  main = "All 2000 genes \nfor PC1 and PC2"
)
legend(-45, -25,
  legend = c("Normal tissue", "Tumor tissue"), bty = "n",
  col = c("red", "blue"), pch = c(16, 16), cex = 1
)
plot(SPC1, SPC2,
  col = cols[Y], xlab = "SPC1", ylab = "SPC2", pch = 16,
  main = paste(non_zero1, "genes for SPC1 \n and", non_zero2, "genes for SPC2")
)
legend(-45, -25,
  legend = c("Normal tissue", "Tumor tissue"), bty = "n",
  col = c("red", "blue"), pch = c(16, 16), cex = 1
)
```

__Conclusion:__ Only about 
`r sprintf("%.2f", 100 * non_zero1 / length(sparse_loadings1))`% (`r non_zero1`) of the genes are useful for PC1 and only about
`r sprintf("%.2f", 100 * non_zero2 / length(sparse_loadings2))`% (`r non_zero2`) of the genes are useful for PC2 .
Sparse PCA has succeeded in setting the uninformative genes/loadings to zero.
In seperating normal and tumour tissues, SPCA performs vitually the same as PCA.
The key point here is that SPCA uses only a minor proportion of the original features to achieve the same results, suggesting that the largest variability of the data is only driven by a minority of features. 

</details>


# LDA

In this section, we will perform LDA on the gene data to get a clear understanding on the genes responsible for separating the tumor and normal tissue groups.

Remember that the LDA problem can be stated as

$$
\mathbf v=\text{ArgMax}_a\frac{\boldsymbol a^t \mathbf B \boldsymbol a}{\boldsymbol a^t \mathbf W \boldsymbol a} \text{ subject to }
\boldsymbol a^t \mathbf W \boldsymbol a=1.
\$$

Which is equivalent to the eigenvalue/eigenvector problem 

$$
\mathbf W^{-1} \mathbf B \boldsymbol a=\lambda \boldsymbol a
$$

In our case where we only have two groups, only one solution exist. This is the one eigenvector $\boldsymbol v$ and one eigenvalue. We can then write the PC/Scores as

$$
\mathbf Z=\mathbf X \boldsymbol v
$$


## Exercises {-}

#### 1. The function `lda()` in the `MASS` package performs LDA. Similar to the `glmnet()` function, you will need to supply an `x` argument. The argument `grouping` is the vector with the response, and this has to be a factor variable. You have that stored as `Y`. Fit an LDA on `X` with grouping `Y`. {-}

<details><summary>Solution</summary>

```{r}
## Perform LDA
alon_lda <- lda(x = X, grouping = Y)
```

</details>

#### 2. $\boldsymbol v$ can be extracted from the object as the element `scaling`. Extract this and call it `V1`. {-}

<details><summary>Solution</summary>

```{r}
V1 <- alon_lda$scaling
```

</details>

#### 3. Compute $\mathbf Z$ and call it `Z1`. {-}

<details><summary>Solution</summary>

```{r}
Z1 <- X %*% V1
```

</details>

#### 4. Now check to see how well your single LDA/`Z1` separates the tumour and normal tissues groups. Compare it to the plot in (3) of the previous exercise, and observe whether LDA performs better in separating the two groups. {-}

<details><summary>Solution</summary>

```{r}
par(mfrow = c(1, 1))
boxplot(Z1 ~ Y, col = cols, ylab = "LDA")
```

</details>

#### 5. As was the case with the first and second PC, `Z1` is a linear combination determined by the loadings $\boldsymbol v$. These are non-zero for all genes. To get a few interesting genes, you can use a sparse LDA. Note that you can use the package `sparseLDA` with the function `sda()` to perform this analysis, but let's do this as we did for sparse PCA. {-}

a. Use the `cv.glmnet` function with `x=X`, `y=Z1` and `alpha=0.5` to select an appropriate number of non-zero genes for the LDA.

<details><summary>Solution</summary>

```{r}
set.seed(45)
lda_loadings <- cv.glmnet(X, Z1, alpha = 0.5, nfolds = 5)
plot(lda_loadings)
```

</details>

b. Check to see how well this subset of genes does in separating the tumour and normal tissue groups. Are they as effective as the entire set of genes?

<details><summary>Solution</summary>

```{r}
sparse_lda_loadings <- as.vector(
  coef(lda_loadings, s = lda_loadings$lambda.1se)
)

# See the genes involved
plot(sparse_lda_loadings[sparse_lda_loadings != 0],
  pch = 16, type = "n", xlim = c(0, 20)
)
text(
  sparse_lda_loadings[sparse_lda_loadings != 0],
  colnames(X)[sparse_lda_loadings != 0]
)
abline(h = 0, lwd = 3)

# without the intercept
SLDA <- X %*% sparse_lda_loadings[-1]

# number of non-zero loadings
n_nonzero <- sum(sparse_lda_loadings > 0)

# boxplots
par(mfrow = c(1, 2))
boxplot(Z1 ~ Y,
  col = cols, ylab = "LDA",
  main = "Entire set of 2000 genes"
)
boxplot(SLDA ~ Y,
  col = cols, ylab = "SLDA",
  main = sprintf("Subset of %d genes", n_nonzero)
)
```

</details>


# References {-}
